# Prepare data

## Load data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata, warning=FALSE, comments="", message=FALSE}
sample_metadata <- read_tsv("data/metadata.tsv")
```

### Read counts

```{r load_read_counts, warning=FALSE, comments="", message=FALSE}
read_counts_raw <- read_tsv("data/genome.count.tsv") %>%
    rename(genome=1)

#Transformation of read_counts to combine data from both sequence rounds
merge_and_rename <- function(read_counts_raw) {
  read_counts_raw %>%
    # Gather the columns into long format
    pivot_longer(cols = -genome, names_to = "col") %>%
    # Extract prefix
    mutate(prefix = gsub("^(.*?)_.*", "\\1", col)) %>%
    # Group by prefix and genome, then summarize
    group_by(prefix, genome) %>%
    summarize(value = sum(value)) %>%
    # Spread the data back to wide format
    pivot_wider(names_from = prefix, values_from = value)
}

read_counts <- merge_and_rename(read_counts_raw)
```

### Genome base hits

```{r load_genome_hits, warning=FALSE, comments="", message=FALSE}
genome_hits_raw <- read_tsv("data/genome.covered_bases.tsv") %>%
    rename(genome=1)

#Transformation of genome_hits to combine data from both sequence rounds
merge_and_rename <- function(genome_hits_raw) {
  genome_hits_raw %>%
    # Gather the columns into long format
    pivot_longer(cols = -genome, names_to = "col") %>%
    # Extract prefix
    mutate(prefix = gsub("^(.*?)_.*", "\\1", col)) %>%
    # Group by prefix and genome, then summarize
    group_by(prefix, genome) %>%
    summarize(value = sum(value)) %>%
    # Spread the data back to wide format
    pivot_wider(names_from = prefix, values_from = value)
}

genome_hits <- merge_and_rename(genome_hits_raw)
```

### Genome taxonomy

```{r load_genome_taxonomy, warning=FALSE, comments="", message=FALSE}
genome_taxonomy <- read_tsv("data/gtdbtk.summary.tsv") %>%
  select(mag_id = user_genome, classification) %>%
  separate(
    classification,
    into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";") %>%
    rename(genome=1)
```

### Genome quality

```{r load_genome_quality, warning=FALSE, comments="", message=FALSE}
genome_quality <- read_tsv("data/quality_report.tsv") %>%
  select(
    genome = Name, 
    completeness = Completeness, 
    contamination = Contamination,
    length = Genome_Size, 
    gc = GC_Content
  )
genome_quality<-genome_quality %>% 
  mutate (genome=str_remove_all(genome,".fa"))

#Filter MAGs with over 70% completeness and less than 10% contamination
genome_quality <- genome_quality %>%
  filter(completeness > 70 & contamination < 10)
```

### Genome tree

```{r load_genome_tree, warning=FALSE, comments="", message=FALSE}
genome_tree <- read_tree("data/gtdbtk.backbone.bac120.classify.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names

#Filter genome_taxonomy to keep MAGs with over 70% completeness and less than 10% contamination
genome_taxonomy <- genome_taxonomy %>%
  semi_join(genome_quality, by = "genome")
genome_tree <- keep.tip(genome_tree, tip=genome_taxonomy$genome) # keep only MAG tips
```

### Genome annotations

```{r load_genome_annotations, warning=FALSE, comments="", message=FALSE}
genome_annotations <- read_tsv("data/annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)

#Filter only the MAGs with over 70% completeness and less than 10% contamination
genome_annotations <- genome_annotations %>%
  semi_join(genome_quality, by = "genome")
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Merge genome taxonomy and quality

```{r generate_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_metadata <- genome_taxonomy %>%
  inner_join(genome_quality,by=join_by(genome==genome)) #join quality
```

### Calculate genome coverage

```{r calculate_genome_coverage, warning=FALSE, comments="", message=FALSE}
#Filter genome_hits for the MAGs with over 70% completeness and less than 10% contamination
genome_hits <- genome_hits %>%
  semi_join(genome_quality, by = "genome")

genome_coverage <- genome_hits %>%
  mutate(across(where(is.numeric), ~ ./genome_metadata$length))
```

## Filtering

Remove antibiotics and transplant samples

```{r filterging_2, echo=TRUE, warning=FALSE}
#Counts_raw
columns_to_exclude <- c("AD16","AD23", "AD25", "AD91", "AFU87", "AFU88", "AFU91", "AFU92", "AFU93", "AFU94", "AFU98", "AFU95", "AFU99", "AFU96", "AFV02", "AFU97", "AFV03", "AFV04", "AFV01", "AFV05", "AFV06", "AFV07",
                        "AFV08", "AFV09", "AFV10", "AFV11", "AFV12", "AFV14", "AFV15", "AFV16", "AFV17") # Columns to exclude
read_counts <- read_counts %>%
  select(-columns_to_exclude)

#Metadata
sample_metadata <- sample_metadata %>%
  filter(time_point != "0_Wild")%>%
  filter(Tube_code != "AD16") %>%
  filter(Tube_code != "AD23") %>%
  filter(Tube_code != "AD25") %>%
  filter(Tube_code != "AD91")

#Coverage_table
columns_to_exclude <- c("AD16","AD23", "AD25", "AD91", "AFU87", "AFU88", "AFU91", "AFU92", "AFU93", "AFU94", "AFU98", "AFU95", "AFU99", "AFU96", "AFV02", "AFU97", "AFV03", "AFV04", "AFV01", "AFV05", "AFV06", "AFV07",
                        "AFV08", "AFV09", "AFV10", "AFV11", "AFV12", "AFV14", "AFV15", "AFV16", "AFV17")  # Columns to exclude
genome_coverage <- genome_coverage %>%
  select(-columns_to_exclude)
```

```{r filterging_3, echo=TRUE, warning=FALSE}
##Removal of samples of individual LI1_2nd_6
#Counts_raw
columns_to_exclude <- c("AFV13","AD06", "AD31", "AE03", "AF12") # Columns to exclude
read_counts <- read_counts %>%
  select(-columns_to_exclude)

#Metadata
sample_metadata<- sample_metadata %>%
  filter(Tube_code != "AFV13") %>%
  filter(Tube_code != "AD06") %>%
  filter(Tube_code != "AD31") %>%
  filter(Tube_code != "AE03") %>%
  filter(Tube_code != "AF12") 

#Coverage_table
columns_to_exclude <- c("AFV13","AD06", "AD31", "AE03", "AF12")  # Columns to exclude
genome_coverage <- genome_coverage %>%
  select(-columns_to_exclude)
```

### Filter reads by coverage

```{r filter_coverage_1, warning=FALSE, comments="", message=FALSE}
#Filter read_counts for the MAGs with over 70% completeness and less than 10% contamination
read_counts<- read_counts %>%
  semi_join(genome_quality, by = "genome")

min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]])) 
```

### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered_1, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered_1, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

## Load data statistics

### Raw reads

```{r raw_reads_1, warning=FALSE, comments="", message=FALSE}
raw_reads <-
  "data/multiqc_general_stats_2.txt" %>%
  read_tsv() %>%
  select(
    sample = Sample,
    raw_reads = `total_sequences`
  ) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(raw_reads = sum(raw_reads, na.rm = TRUE)) %>%
  rename(sample = sample_prefix)  %>%
  mutate(sample = str_remove(sample, "^fastp \\|\\s*")) %>%
  filter(sample!="AD16") %>%
  filter(sample!="AFV08") %>%
  filter(sample!="AD23") %>%
  filter(sample!= "AD25") %>%
  filter(sample!= "AD91") %>%
  filter(sample!= "AFU87") %>%
  filter(sample!= "AFU88") %>%
  filter(sample!= "AFU91") %>%
  filter(sample!= "AFU92") %>%
  filter(sample!= "AFU93") %>%
  filter(sample!= "AFU94") %>%
  filter(sample!= "AFU98") %>%
  filter(sample!= "AFU95") %>%
  filter(sample!= "AFU99") %>%
  filter(sample!= "AFU96") %>%
  filter(sample!= "AFV02") %>%
  filter(sample!= "AFU97") %>%
  filter(sample!= "AFV03") %>%
  filter(sample!= "AFV04") %>%
  filter(sample!= "AFV01") %>%
  filter(sample!= "AFV05") %>%
  filter(sample!= "AFV06") %>%
  filter(sample!= "AFV07") %>%
  filter(sample!= "AFV09") %>%
  filter(sample!= "AFV10") %>%
  filter(sample!= "AFV11") %>%
  filter(sample!= "AFV12") %>%
  filter(sample!= "AFV14") %>%
  filter(sample!= "AFV15") %>%
  filter(sample!= "AFV16") %>%
  filter(sample!= "AFV17") %>%
  filter(sample!= "AFV13")
```

### Quality-filtered reads

```{r filtered_reads_1, warning=FALSE, comments="", message=FALSE}
fastp_reads <-
  "data/multiqc_general_stats.txt" %>%
  read_tsv() %>%
  select(
    sample = Sample,
    trimmed_reads = `total_sequences`
  ) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(trimmed_reads = sum(trimmed_reads, na.rm = TRUE)) %>%
  rename(sample = sample_prefix) %>% 
  filter(!str_detect(sample, "nonlizard \\|")) %>%
  filter(!str_detect(sample, "lizard \\|")) %>%
  filter(!str_detect(sample, "refseq500 \\|"))  %>%
  mutate(sample = str_remove(sample, "^fastp \\|\\s*")) %>%
  filter(sample!="AD16") %>%
  filter(sample!="AFV08") %>%
  filter(sample!="AD23") %>%
  filter(sample!= "AD25") %>%
  filter(sample!= "AD91") %>%
  filter(sample!= "AFU87") %>%
  filter(sample!= "AFU88") %>%
  filter(sample!= "AFU91") %>%
  filter(sample!= "AFU92") %>%
  filter(sample!= "AFU93") %>%
  filter(sample!= "AFU94") %>%
  filter(sample!= "AFU98") %>%
  filter(sample!= "AFU95") %>%
  filter(sample!= "AFU99") %>%
  filter(sample!= "AFU96") %>%
  filter(sample!= "AFV02") %>%
  filter(sample!= "AFU97") %>%
  filter(sample!= "AFV03") %>%
  filter(sample!= "AFV04") %>%
  filter(sample!= "AFV01") %>%
  filter(sample!= "AFV05") %>%
  filter(sample!= "AFV06") %>%
  filter(sample!= "AFV07") %>%
  filter(sample!= "AFV09") %>%
  filter(sample!= "AFV10") %>%
  filter(sample!= "AFV11") %>%
  filter(sample!= "AFV12") %>%
  filter(sample!= "AFV14") %>%
  filter(sample!= "AFV15") %>%
  filter(sample!= "AFV16") %>%
  filter(sample!= "AFV17") %>%
  filter(sample!= "AFV13")
```

### Host-mapped reads

```{r host_mapped_1, warning=FALSE, comments="", message=FALSE}
host_mapped <-
  "data/multiqc_general_stats.txt" %>%
  read_tsv() %>%
  filter(str_detect(Sample, "lizard")) %>%
  select(
    sample = Sample,
    host_mapped = `reads_mapped`,
    mapping_total = `raw_total_sequences`
  ) %>%
  mutate(
    host_unmapped = mapping_total - host_mapped
  ) %>%
  filter(!is.na(host_mapped)) %>%
  separate(
    col = sample,
    into = c("host_name", "sample"),
    sep = " \\| "
  ) %>%
  rename(mapped = host_mapped, unmapped = host_unmapped) %>%
  select(-mapping_total) %>%
  pivot_longer(-host_name:-sample) %>%
  mutate(
    name = str_glue("{name}_{host_name}")
  ) %>%
  select(-host_name) %>%
  pivot_wider() %>%
  mutate(sample = str_extract(sample, "^[^_]+")) %>%
  group_by(sample) %>%
  summarize(
    mapped_lizard = sum(mapped_lizard),
    unmapped_lizard = sum(unmapped_lizard)) %>%
  filter(sample!="AD16") %>%
  filter(sample!="AFV08") %>%
  filter(sample!="AD23") %>%
  filter(sample!= "AD25") %>%
  filter(sample!= "AD91") %>%
  filter(sample!= "AFU87") %>%
  filter(sample!= "AFU88") %>%
  filter(sample!= "AFU91") %>%
  filter(sample!= "AFU92") %>%
  filter(sample!= "AFU93") %>%
  filter(sample!= "AFU94") %>%
  filter(sample!= "AFU98") %>%
  filter(sample!= "AFU95") %>%
  filter(sample!= "AFU99") %>%
  filter(sample!= "AFU96") %>%
  filter(sample!= "AFV02") %>%
  filter(sample!= "AFU97") %>%
  filter(sample!= "AFV03") %>%
  filter(sample!= "AFV04") %>%
  filter(sample!= "AFV01") %>%
  filter(sample!= "AFV05") %>%
  filter(sample!= "AFV06") %>%
  filter(sample!= "AFV07") %>%
  filter(sample!= "AFV09") %>%
  filter(sample!= "AFV10") %>%
  filter(sample!= "AFV11") %>%
  filter(sample!= "AFV12") %>%
  filter(sample!= "AFV14") %>%
  filter(sample!= "AFV15") %>%
  filter(sample!= "AFV16") %>%
  filter(sample!= "AFV17") %>%
  filter(sample!= "AFV13")
```

### Prokaryotic fraction

```{r singlem_1, warning=FALSE, comments="", message=FALSE}
singlem <-
  "data/singleM.tsv" %>%
  read_tsv() %>%
  distinct() %>%
  mutate(
    sample = sample %>% str_remove_all("_1$"),
   read_fraction = read_fraction %>% str_remove("%") %>% as.numeric(),
   read_fraction = read_fraction / 100
  ) %>%
  select(
   sample,
   singlem_prokaryotic_bases = bacterial_archaeal_bases,
   singlem_metagenome_size = metagenome_size,
   singlem_read_fraction = read_fraction,
  ) %>%
mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(
    singlem_prokaryotic_bases = sum(singlem_prokaryotic_bases),
    singlem_metagenome_size = sum(singlem_metagenome_size),
    singlem_read_fraction = mean(singlem_read_fraction)
  ) %>%
  rename(sample = sample_prefix)%>%
   filter(sample!="AD16") %>%
  filter(sample!="AFV08") %>%
  filter(sample!="AD23") %>%
  filter(sample!= "AD25") %>%
  filter(sample!= "AD91") %>%
  filter(sample!= "AFU87") %>%
  filter(sample!= "AFU88") %>%
  filter(sample!= "AFU91") %>%
  filter(sample!= "AFU92") %>%
  filter(sample!= "AFU93") %>%
  filter(sample!= "AFU94") %>%
  filter(sample!= "AFU98") %>%
  filter(sample!= "AFU95") %>%
  filter(sample!= "AFU99") %>%
  filter(sample!= "AFU96") %>%
  filter(sample!= "AFV02") %>%
  filter(sample!= "AFU97") %>%
  filter(sample!= "AFV03") %>%
  filter(sample!= "AFV04") %>%
  filter(sample!= "AFV01") %>%
  filter(sample!= "AFV05") %>%
  filter(sample!= "AFV06") %>%
  filter(sample!= "AFV07") %>%
  filter(sample!= "AFV09") %>%
  filter(sample!= "AFV10") %>%
  filter(sample!= "AFV11") %>%
  filter(sample!= "AFV12") %>%
  filter(sample!= "AFV14") %>%
  filter(sample!= "AFV15") %>%
  filter(sample!= "AFV16") %>%
  filter(sample!= "AFV17") %>%
  filter(sample!= "AFV13")
```

### MAG-mapped reads

```{r mag_mapping_1, warning=FALSE, comments="", message=FALSE}
mag_mapping <-
  read_tsv("data/contig.count.tsv.xz") %>%
  pivot_longer(-sequence_id) %>%
  summarise(value = sum(value), .by = "name") %>%
  rename(sample = name, mapped_mags = value) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
    group_by(sample_prefix) %>%
    summarize(
      mapped_mags = sum(mapped_mags)
    ) %>%
    rename(sample = sample_prefix)%>%
  filter(sample!="AD16") %>%
  filter(sample!="AFV08") %>%
  filter(sample!="AD23") %>%
  filter(sample!= "AD25") %>%
  filter(sample!= "AD91") %>%
  filter(sample!= "AFU87") %>%
  filter(sample!= "AFU88") %>%
  filter(sample!= "AFU91") %>%
  filter(sample!= "AFU92") %>%
  filter(sample!= "AFU93") %>%
  filter(sample!= "AFU94") %>%
  filter(sample!= "AFU98") %>%
  filter(sample!= "AFU95") %>%
  filter(sample!= "AFU99") %>%
  filter(sample!= "AFU96") %>%
  filter(sample!= "AFV02") %>%
  filter(sample!= "AFU97") %>%
  filter(sample!= "AFV03") %>%
  filter(sample!= "AFV04") %>%
  filter(sample!= "AFV01") %>%
  filter(sample!= "AFV05") %>%
  filter(sample!= "AFV06") %>%
  filter(sample!= "AFV07") %>%
  filter(sample!= "AFV09") %>%
  filter(sample!= "AFV10") %>%
  filter(sample!= "AFV11") %>%
  filter(sample!= "AFV12") %>%
  filter(sample!= "AFV14") %>%
  filter(sample!= "AFV15") %>%
  filter(sample!= "AFV16") %>%
  filter(sample!= "AFV17") %>%
  filter(sample!= "AFV13")
```

### Wrap data statistics

```{r wrap_statistics_1, warning=FALSE, comments="", message=FALSE}
data_stats <- raw_reads %>%
  left_join(fastp_reads) %>%
  left_join(host_mapped) %>%
  left_join(singlem) %>%
  left_join(mag_mapping)

data_stats<- data_stats %>%
  filter(!str_detect(sample, "nonlizard \\|")) %>%
  filter(!str_detect(sample, "lizard \\|")) %>%
  filter(!str_detect(sample, "refseq500 \\|"))
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects_1, warning=FALSE, comments="", message=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     phylum_colors,
     data_stats,
     file = "data/data.Rdata")
```
