# Are the microbial communities similar in both donor samples?
```{r}
load("data/objects_2003.Rdata")
load("data/beta_08032025.Rdata")
load("data/alpha_08032025.Rdata")
```
## Alpha diversity

```{r filter_data_donor, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type =="Hot_control" & time_point %in% c( "Transplant1", "Transplant2"))
```

```{r filter_Data2, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type =="Hot_control" & time_point %in% c( "Transplant1", "Transplant2"))%>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type =="Hot_control" & time_point %in% c( "Transplant1", "Transplant2"))
```

```{r alpha_div_plot_02, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type =="Hot_control" & time_point %in% c( "Transplant1", "Transplant2")) %>% 
  ggplot(aes(y = value, x = time_point, color=time_point, fill=time_point)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha=0.5) +
  scale_color_manual(name="time_point",
          breaks=c("Transplant1","Transplant2"),
          values=c('#d5992c', "#d5b52c")) +
      scale_fill_manual(name="time_point",
          breaks=c("Transplant1","Transplant2"),
          values=c('#d5992c50', "#d5b52c50")) +
  facet_wrap(. ~ metric, scales = "free") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    # Increase plot size
    plot.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 8)
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_donor, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ time_point+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_donor, comment="", message=FALSE, warning=FALSE}
Model_neutral <- nlme::lme(fixed = neutral ~ time_point, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_donor, comment="", message=FALSE, warning=FALSE}
Model_phylo <-  nlme::lme(fixed = phylogenetic ~ time_point, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity

***Richness***
```{r permanova_richness2_donor, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.)
```
```{r permanova_richness2_donor_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral2, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral2_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(Tube_code,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo2, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_phylo2_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(Tube_code,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

# Does the donor sample maintain the microbial community found in acclimation?

```{r filter_data_donor2, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
    mutate(treatment=case_when(
    treatment %in% c("Transplant1", "Transplant2") ~ "Transplant",
    TRUE ~ treatment
  ))

alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Hot_control" & treatment %in% c("Acclimation","Transplant"))
```

```{r filter_data3, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
    mutate(treatment=case_when(
    treatment %in% c("Transplant1", "Transplant2") ~ "Transplant",
    TRUE ~ treatment
  ))
samples_to_keep <- sample_metadata %>%
  filter(type == "Hot_control" & treatment %in% c("Acclimation","Transplant")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Hot_control" & treatment %in% c("Acclimation","Transplant"))
```

## Alpha diversity

```{r alpha_div_plot_002, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type == "Hot_control" & treatment %in% c("Acclimation","Transplant")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha=0.5) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c("#d57d2c", "#d5b52c")) +
      scale_fill_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c("#d57d2c50", "#d5b52c50")) +
  facet_wrap(. ~ metric, scales = "free") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    # Increase plot size
    plot.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 8)
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_fmt, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_fmt, comment="", message=FALSE, warning=FALSE}
Model_neutral <- nlme::lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_fmt, comment="", message=FALSE, warning=FALSE}
Model_phylo <- nlme::lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```


## Beta diversity
***Richness***
```{r permanova_richness3, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_richness3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral3, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo3, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_phylo3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE}
p0<-richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c("#d57d2c", "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

p1<-neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c("#d57d2c", "#d5b52c")) +  
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

p2<-phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c("#d57d2c", "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```

```{r beta_neutral_nmds_plot_donors_final, fig.height=8, fig.width=12, fig.fullwidth=TRUE}
ggarrange(p0, p1, p2, ncol=3, nrow=1, common.legend = TRUE, legend="right")
```

# Is the donor sample microbiota different to recipient microbiota?



## Alpha diversity

```{r filter_data_donor3, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
    mutate(treatment=case_when(
    treatment %in% c("Transplant1", "Transplant2") ~ "Transplant",
    TRUE ~ treatment
  ))

alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Treatment" & treatment %in% c("Acclimation","Transplant"))
```

```{r filter_data4, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Treatment" & treatment %in% c("Acclimation", "Transplant")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Treatment" & treatment %in% c("Acclimation", "Transplant"))
```

```{r alpha_div_plot_03, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type == "Treatment" & treatment %in% c("Acclimation", "Transplant")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha=0.5) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c('#008080', "#d5b52c")) +
      scale_fill_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c('#00808050', "#d5b52c50")) +
  facet_wrap(. ~ metric, scales = "free") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    # Increase plot size
    plot.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 8)
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_fmt1, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_fmt1, comment="", message=FALSE, warning=FALSE}
Model_neutral <-nlme:: lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_fmt1, comment="", message=FALSE, warning=FALSE}
Model_phylo <- nlme::lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity
***Richness***
```{r permanova_richness4, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_richness4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral4, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo4, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_phylo4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

```{r beta_div_nmds_1_plot, comment="", message=FALSE, warning=FALSE}
p0<-richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

p1<-neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c('#008080', "#d5b52c")) +  
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

p2<-phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="treatment",
          breaks=c("Acclimation","Transplant"),
          values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```

```{r beta_neutral_nmds_plot_donors_1_final, fig.height=8, fig.width=12, fig.fullwidth=TRUE}
ggarrange(p0, p1, p2, ncol=3, nrow=1, common.legend = TRUE, legend="right")
```