# Are the microbial communities similar in both donor samples?
```{r data_donor, comment="", message=FALSE, warning=FALSE, echo=FALSE}
load("data/objects_2803.Rdata")
load("data/beta_08032025.Rdata")
load("data/alpha_08032025.Rdata")
```

## Alpha diversity
```{r filter_data_donor, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type =="Warm_control" & time_point %in% c( "Donor1", "Donor2"))
```

```{r filter_Data2, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type =="Warm_control" & time_point %in% c( "Donor1", "Donor2"))%>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type =="Warm_control" & time_point %in% c( "Donor1", "Donor2"))
```

```{r alpha_div_plot_02, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type =="Warm_control" & time_point %in% c( "Donor1", "Donor2")) %>% 
  ggplot(aes(y = value, x = time_point, color=time_point, fill=time_point)) +
  geom_boxplot(width = 0.5,alpha=0.5, outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.2,show.legend = FALSE) +
  scale_color_manual(name="Time point",values=c('#d5992c', "#d5b52c")) +
  scale_fill_manual(name="Time point",values=c('#d5992c', "#d5b52c"))  +
  facet_wrap(. ~ metric, scales = "free") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 18, lineheight = 0.6),
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_donor, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ time_point+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_donor, comment="", message=FALSE, warning=FALSE}
Model_neutral <- nlme::lme(fixed = neutral ~ time_point, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_donor, comment="", message=FALSE, warning=FALSE}
Model_phylo <-  nlme::lme(fixed = phylogenetic ~ time_point, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity

***Richness***
```{r permanova_richness2_donor, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$time_point) %>% permutest(.)
```
```{r permanova_richness2_donor_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ time_point,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral2, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$time_point) %>% permutest(.) 
```
```{r permanova_neutral2_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ time_point,
        data = subset_meta %>% arrange(match(Tube_code,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo2, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$time_point) %>% permutest(.) 
```
```{r permanova_phylo2_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ time_point,
        data = subset_meta %>% arrange(match(Tube_code,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

# Does the donor sample maintain the microbial community found in acclimation?
```{r filter_data_donor2, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
  mutate(treatment=time_point) %>% 
    mutate(treatment=case_when(
    treatment %in% c("Donor1", "Donor2") ~ "Donor",
    TRUE ~ treatment
  ))

alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Warm_control" & treatment %in% c("Acclimation","Donor"))
```

```{r filter_data3, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Warm_control" & treatment %in% c("Acclimation","Donor")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Warm_control" & treatment %in% c("Acclimation","Donor"))
```

## Alpha diversity
```{r alpha_div_plot_002, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type == "Warm_control" & treatment %in% c("Acclimation","Donor")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
  geom_boxplot(width = 0.5,alpha=0.5 ,outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.2, show.legend = FALSE) +
  scale_color_manual(values=c('#008080', "#d5b52c")) +
  scale_fill_manual(values=c('#008080', "#d5b52c")) +
  facet_wrap( ~ metric, scales = "free")+
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.text = element_text(size = 12),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_blank(),
    strip.text = element_text(size = 16, lineheight = 0.6)
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_fmt, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_fmt, comment="", message=FALSE, warning=FALSE}
Model_neutral <- nlme::lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_fmt, comment="", message=FALSE, warning=FALSE}
Model_phylo <- nlme::lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```


## Beta diversity

***Richness***
```{r permanova_richness3, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_richness3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral3, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo3, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_phylo3_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***NMDS***
```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE}
richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.6, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.6, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.6, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```

# Is the donor sample microbiota different to recipient microbiota?

## Alpha diversity

```{r filter_data_donor3, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation","Donor"))
```

```{r filter_data4, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Donor")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Donor"))
```

```{r alpha_div_plot_03, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Donor")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
    geom_boxplot(width = 0.5,alpha=0.5 ,outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.2, show.legend = FALSE) +
  scale_color_manual(values=c('#008080', "#d5b52c")) +
  scale_fill_manual(values=c('#008080', "#d5b52c")) +
  facet_wrap( ~ metric, scales = "free")+
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.text = element_text(size = 12),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_blank(),
    strip.text = element_text(size = 16, lineheight = 0.6)
      ) +
  ylab("Alpha diversity")
```

***Richness***
```{r LMM_fmt1, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral_fmt1, comment="", message=FALSE, warning=FALSE}
Model_neutral <-nlme:: lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylol_fmt1, comment="", message=FALSE, warning=FALSE}
Model_phylo <- nlme::lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity
***Richness***
```{r permanova_richness4, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_richness4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral4, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo4, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_phylo4_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(Tube_code,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***NMDS***
```{r beta_div_nmds_1_plot, comment="", message=FALSE, warning=FALSE}
richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c('#008080', "#d5b52c")) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```