# Does FMT change the microbial community over time?
```{r data_FMT, comment="", message=FALSE, warning=FALSE, echo=FALSE}
load("data/objects_2803.Rdata")
load("data/beta_08032025.Rdata")
load("data/alpha_08032025.Rdata")
```

## Alpha diversity
```{r filter_data5, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
  mutate(treatment=time_point) %>% 
    mutate(treatment=case_when(
    treatment %in% c("Donor1", "Donor2") ~ "Donor",
    TRUE ~ treatment
  ))

alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1","FMT2") ) 
```

```{r alpha_div_plot_5, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral", "phylogenetic"))) %>%
  filter(type=="Cold_intervention" & treatment %in% c("FMT1", "FMT2")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
  geom_boxplot(width = 0.5,alpha=0.5 ,outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.2, show.legend = FALSE) +
  scale_color_manual(values=c("#76b183", '#40714b50')) +
  scale_fill_manual(values=c("#76b183", '#40714b50')) +
  facet_wrap( ~ metric, scales = "free")+
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.text = element_text(size = 12),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_blank(),
    strip.text = element_text(size = 16, lineheight = 0.6)
      )  +
  ylab("Alpha diversity")

```

***Richness***
```{r LMM_rich5, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral5, comment="", message=FALSE, warning=FALSE}
Model_neutral <- nlme::lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylo5, comment="", message=FALSE, warning=FALSE}
Model_phylo <- nlme::lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity
```{r filter_data_beta5, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1", "FMT2")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1", "FMT2"))
```

***Richness***
```{r permanova_rich5, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_rich5_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Neutral***
```{r permanova_neutral5, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_neutral5_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***Phylogenetic***
```{r permanova_phylo5, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_phylo5_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

***NMDS***
```{r beta_div_nmds_2_plot, comment="", message=FALSE, warning=FALSE}
richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c("#76b183", '#40714b50')) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c("#76b183", '#40714b50')) +  
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(name="Time point",values=c("#76b183", '#40714b50')) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```


# Do FMT change the microbial community compared to antibiotics and acclimation?
```{r filter_data61, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Cold_intervention" & treatment %in% c("Antibiotics", "Acclimation","FMT1","FMT2") ) 
```

## Alpha diversity

***Richness*** 
```{r LMM_rich6, comment="", message=FALSE, warning=FALSE}
# Modelq0GLMMNB <- glmer.nb(richness ~ treatment + (1|individual), data = alpha_div_meta)
# summary(Modelq0GLMMNB)
#emmeans(Modelq0GLMMNB, pairwise ~ treatment)

model_nb <- MASS::glm.nb(richness ~ treatment, data = alpha_div_meta)
summary(model_nb)
emmeans(model_nb, pairwise ~ treatment)

```

***Neutral***
```{r filter_neutral_data61, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
emmeans(Model_neutral, pairwise ~ treatment)
```

```{r LMM_neutral6, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
emmeans(Model_neutral, pairwise ~ treatment)
```

***Phylogenetic***
```{r LMM_phylo6, comment="", message=FALSE, warning=FALSE}
Model_phylo <- lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
emmeans(Model_phylo, pairwise ~ treatment)
```

## Beta diversity
```{r filter_data_beta6, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Antibiotics", "FMT1", "FMT2")) %>% 
  select(sample) %>% 
  pull()
subset_meta <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Antibiotics", "FMT1", "FMT2"))
```

***Richness***
```{r permanova_rich6, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_rich6_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```
```{r pairwise_rich6, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(richness, subset_meta$treatment, perm = 999)
```

***Neutral***
```{r permanova_neutral6, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.)
```

```{r permanova_neutral6_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```
```{r pairwise_neutral6, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(neutral, subset_meta$treatment, perm = 999)
```

***Phylogenetic***
```{r permanova_phylo6, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_phylo6_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```
```{r pairwise_phylo6, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(phylo, subset_meta$treatment, perm = 999)
```


# Is the gut microbiota similar to the donor after FMT?

## Alpha diversity
```{r alpha_div_plot_7, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation", "Antibiotics", "Donor", "FMT1", "FMT2") )

alpha_div_meta$treatment <- factor(alpha_div_meta$treatment, levels=c("Acclimation","Antibiotics","Donor", "FMT1", "FMT2"))

alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = "sample") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  filter(type=="Cold_intervention" & treatment %in% c("Acclimation", "Antibiotics", "Donor", "FMT1", "FMT2")) %>% 
  ggplot(aes(y = value, x = treatment, color=treatment, fill=treatment)) +
  geom_jitter(width = 0.2, show.legend = FALSE) +
  geom_boxplot(width = 0.5, alpha=0.5, outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(name="Time point",values=c('#008080',"#003a45", "#d5b52c", "#76b183",'#40714b50')) +
  scale_fill_manual(name="Time point",values=c('#008080',"#003a45", "#d5b52c", "#76b183",'#40714b50')) +
  facet_wrap(. ~ metric, scales = "free") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_blank(),
    strip.text = element_text(size = 16, lineheight = 0.6)
      )  +
  ylab("Alpha diversity")
```

```{r filter_data_cold|_int, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  filter(type == "Cold_intervention" & treatment %in% c("Donor", "FMT1", "FMT2") )
```


***Richness***
```{r LMM_rich7, comment="", message=FALSE, warning=FALSE}
Modelq0GLMMNB <- glmer.nb(richness ~ treatment+(1|individual), data = alpha_div_meta)
summary(Modelq0GLMMNB)
```

***Neutral***
```{r LMM_neutral7, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lme(fixed = neutral ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_neutral)
```

***Phylogenetic***
```{r LMM_phylo7, comment="", message=FALSE, warning=FALSE}
Model_phylo <- lme(fixed = phylogenetic ~ treatment, data = alpha_div_meta,
               random = ~ 1 | individual)
summary(Model_phylo)
```

## Beta diversity
```{r filter_Data7_beta, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Donor", "FMT1", "FMT2") ) %>% 
  select(sample) %>%
  pull()

subset_meta <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Donor", "FMT1", "FMT2") )
```

***Richness***
```{r permanova_rich7, comment="", message=FALSE, warning=FALSE}
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])

betadisper(richness, subset_meta$treatment) %>% permutest(.) 
```
```{r permanova_rich7_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(richness ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(richness))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(richness))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

```{r pairwise_rich_fmt, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(richness, subset_meta$treatment, perm = 999)
```

***Neutral***
```{r permanova_neutral7, comment="", message=FALSE, warning=FALSE}
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])

betadisper(neutral, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_neutral7_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(neutral ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(neutral))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(neutral))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

```{r pairwise_neutral_fmt, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(neutral, subset_meta$treatment, perm = 999)
```

***Phylogenetic***
```{r permanova_phylo7, comment="", message=FALSE, warning=FALSE}
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
betadisper(phylo, subset_meta$treatment) %>% permutest(.) 
```

```{r permanova_phylo7_adonis, comment="", message=FALSE, warning=FALSE}
adonis2(phylo ~ treatment,
        data = subset_meta %>% arrange(match(sample,labels(phylo))),
        permutations = 999,
        strata = subset_meta %>% arrange(match(sample,labels(phylo))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

```{r pairwise_phylo_fmt, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(phylo, subset_meta$treatment, perm = 999)
```

***NMDS***
```{r filter_Data_nmds, comment="", message=FALSE, warning=FALSE}
samples_to_keep <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation","Antibiotics","Donor", "FMT1", "FMT2") ) %>% 
  select(sample) %>%
  pull()

subset_meta <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("Acclimation","Antibiotics","Donor", "FMT1", "FMT2") )

alpha_div_meta$treatment <- factor(alpha_div_meta$treatment, levels=c("Acclimation","Antibiotics","Donor", "FMT1", "FMT2"))
richness <- as.matrix(beta_q0n$S)
richness <- as.dist(richness[rownames(richness) %in% samples_to_keep,
               colnames(richness) %in% samples_to_keep])
neutral <- as.matrix(beta_q1n$S)
neutral <- as.dist(neutral[rownames(neutral) %in% samples_to_keep,
               colnames(neutral) %in% samples_to_keep])
phylo <- as.matrix(beta_q1p$S)
phylo <- as.dist(phylo[rownames(phylo) %in% samples_to_keep,
               colnames(phylo) %in% samples_to_keep])
```



```{r beta_div_nmds_7_plot, comment="", message=FALSE, warning=FALSE}
richness %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080',"#003a45", "#d5b52c", "#76b183",'#40714b50'))) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

neutral %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080',"#003a45", "#d5b52c", "#76b183",'#40714b50'))) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )

phylo %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(subset_meta, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(name="Time point",values=c('#008080',"#003a45", "#d5b52c", "#76b183",'#40714b50'))) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right", legend.box = "vertical"
    )
```

### Differences in the functional capacities
```{r gift_all_three, comment="", message=FALSE, warning=FALSE}
sample_sub <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1","FMT2", "Donor"))

genome_counts_filtered <- genome_counts_filt %>% 
    select(one_of(c("genome",sample_sub$sample)))

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filtered$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_filtered_row <- genome_counts_filtered %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 

GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_filtered_row,GIFT_db)
```

```{r gitfs_func_fmt_three, comment="", message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  summarise(MCI = mean(value), sd = sd(value))

MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(MCI$value)
kruskal.test(value ~ treatment, data=MCI)
```

## Donor vs FMT1

### Structural zeros
```{r struct_zero_fmt_tranp, comment="", message=FALSE, warning=FALSE}
FMT_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment == "FMT1") %>% 
  dplyr::select(sample) %>%
  pull()

Donor_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment =="Donor") %>% 
  dplyr::select(sample) %>% pull()

sample_sub <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1", "Donor"))

structural_zeros <- genome_counts_filt %>% 
  select(one_of(c("genome",subset_meta$sample))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>% 
   rowwise() %>% #compute for each row (genome)
   mutate(all_zeros_FMT = all(c_across(all_of(FMT_samples)) == 0)) %>% 
   mutate(all_zeros_Donor = all(c_across(all_of(Donor_samples)) == 0)) %>% 
   mutate(average_FMT = mean(c_across(all_of(FMT_samples)), na.rm = TRUE)) %>% 
   mutate(average_Donor = mean(c_across(all_of(Donor_samples)), na.rm = TRUE)) %>% 
   filter(all_zeros_FMT == TRUE || all_zeros_Donor==TRUE)  %>% 
   mutate(present = case_when(
      all_zeros_FMT & !all_zeros_Donor ~ "Donor",
      !all_zeros_FMT & all_zeros_Donor ~ "FMT1",
      !all_zeros_FMT & !all_zeros_Donor ~ "None",
      TRUE ~ NA_character_
    )) %>%
   mutate(average = ifelse(present == "FMT1", average_FMT, average_Donor)) %>%
   dplyr::select(genome, present, average) %>%
   left_join(genome_metadata, by=join_by(genome==genome)) %>%
   arrange(present,-average)
```

***Structural zeros in each group***
```{r struc_count_fmt1_tranp, comment="", message=FALSE, warning=FALSE}
fmt <- structural_zeros %>% 
  filter(present=="FMT1") %>% 
  count(phylum, name = "FMT1") %>%
  arrange(desc(FMT1)) 

fmt_f <- structural_zeros %>% 
  filter(present=="FMT1") %>% 
  count(family, name = "FMT1") %>%
  arrange(desc(FMT1)) 
```

```{r total_sum_struc_tranp_fmt1, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt, by="phylum" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%
  as.data.frame() %>% 
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))
```

***Phyla to which the structural zeros belong in each group***
```{r struc_count_merged_fmt1_tranp, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt, by="phylum" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  tt()
```

***Families to which the structural zeros belong in each group***
```{r total_sum_family_fmt_tranp, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(family, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt_f, by="family" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  tt()
```

#### Functional capacities of the structural zeros
```{r gift_struc_fmt_tranp, comment="", message=FALSE, warning=FALSE}
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% structural_zeros$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  filter(genome %in% structural_zeros$genome) %>% 
  select(one_of(c("genome",sample_sub$sample))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>% 
  column_to_rownames(., "genome")

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

```{r comunity_elem_FMT_Tras, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c(9,10)], by=join_by("sample"=="sample"))
```

```{r commun_wilcox_elem_FMT_trans, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  rownames_to_column(., "Elements")  %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))

element_gift_t <- element_gift  %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "trait")

element_gift_filt <- subset(element_gift_t, trait %in% significant_elements$trait) %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "sample")%>% 
  left_join(., sample_metadata[c(9,10)], by = join_by(sample == sample))

difference_table <- element_gift_filt %>%
  select(-sample) %>%
  group_by(treatment) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=FMT1-Donor)%>% 
  mutate(group_color = ifelse(Difference <0, "Donor","FMT1")) 
```

```{r commun_wilcox_elem_plot_fmt_tranp, comment="", message=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.fullwidth=TRUE}
difference_table %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
  scale_fill_manual(
          breaks=c("FMT1","Donor"),
          values=c('#40714b', "#d5b52c")) +
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Microbial Functional Capacity") + 
  ylab("Mean difference")
```

### Differential abundance analysis: composition
```{r phyloseq_fmt_tranp, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1", "Donor")) %>% 
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>% 
  filter(!genome %in% structural_zeros$genome) %>% 
  select(one_of(c("genome",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
```

```{r ancom_rand_fmt_tranp, comment="", message=FALSE, warning=FALSE}
ancombc_FMT_Donor_2803 = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "treatment",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

```{r save_ancombc_fmt_tranp, warning=FALSE, comments="", message=FALSE}
save(ancombc_FMT_Donor_2803, file = "data/ancombc_FMT_Donor_2803.Rdata")
```

```{r load_ancombc_fmt_tranp, warning=FALSE, comments="", message=FALSE, eval=FALSE}
load("data/ancombc_FMT_Donor_2803.Rdata")
```

```{r ancom_rand_fmt1, comment="", message=FALSE, warning=FALSE}
taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))

ancombc_rand_table_mag <- ancombc_FMT_Donor_2803$res %>%
  dplyr::select(taxon, lfc_treatmentFMT1, p_treatmentFMT1) %>%
  filter(p_treatmentFMT1 < 0.05) %>%
  dplyr::arrange(p_treatmentFMT1) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_treatmentFMT1)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

***Differential abundance MAGs in each group***
```{r diff_fmt_tran, comment="", message=FALSE, warning=FALSE}
fmt_count <- ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT1<0)  %>% 
  count(phylum, name = "FMT1") %>%
  arrange(desc(FMT1)) 

ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT1>0)  %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor))  %>% 
  full_join(.,fmt_count, by="phylum")%>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
```

```{r total_sum_fmt_tran, comment="", message=FALSE, warning=FALSE}
ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT1>0)  %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor))  %>% 
  full_join(.,fmt_count, by="phylum")%>%
  mutate(across(where(is.numeric), ~ replace_na(., 0))) %>% 
  as.data.frame() %>% 
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))
```

```{r ancombc_rand_plot_phy_fmt_tran, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag%>%
      mutate(genome=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
ggplot(., aes(x=lfc_treatmentFMT1, y=forcats::fct_reorder(genome,lfc_treatmentFMT1), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```
MAGs enriched in donor on the left side and in FMT1 on the right side


### Differences in the functional capacities
```{r gift_all_fmt, comment="", message=FALSE, warning=FALSE}
sample_sub <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT1", "Donor"))
genome_counts_filtered <- genome_counts_filt %>% 
    select(one_of(c("genome",sample_sub$sample)))

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filtered$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_filtered_row <- genome_counts_filtered %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_filtered_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_filtered_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_filtered_row,GIFT_db)
```

#### Community elements differences:

***MCI***
```{r gitfs_elem_fmt_tran, comment="", message=FALSE, warning=FALSE}
GIFTs_elements_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  summarise(MCI = mean(value), sd = sd(value))

MCI <- GIFTs_elements_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(MCI$value)
wilcox.test(value ~ treatment, data=MCI)
```
```{r comunity_elem_fmt_tran, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c(9,10)], by=join_by("sample"=="sample"))
```

```{r commun_wilcox_elem_fmt_tran, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  rownames_to_column(., "Elements")  %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))

element_gift_t <- element_gift  %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "trait")

element_gift_filt <- subset(element_gift_t, trait %in% significant_elements$trait) %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "sample")%>% 
  left_join(., sample_metadata[c(9,10)], by = join_by(sample == sample))

difference_table <- element_gift_filt %>%
  select(-sample) %>%
  group_by(treatment) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=FMT1-Donor)%>% 
  mutate(group_color = ifelse(Difference <0, "Donor","FMT1")) 
```

#### Community functions differences

***MCI***
```{r gitfs_func_fmt_tran, comment="", message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  summarise(MCI = mean(value), sd = sd(value))

MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(MCI$value)
wilcox.test(value ~ treatment, data=MCI)
```
```{r comunity_func_fmt_tran, comment="", message=FALSE, warning=FALSE}
function_gift <- GIFTs_functions_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  merge(., sample_metadata[c(9,10)], by="sample")
```

```{r commun_wilcox_func_fmt_tran, comment="", message=FALSE, warning=FALSE}
unique_funct_db<- GIFT_db[c(3,4,5)] %>% 
  distinct(Code_function, .keep_all = TRUE)

significant_functional <- function_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  left_join(.,unique_funct_db[c(1,3)],by = join_by(trait == Code_function))

significant_functional
```

## Donor vs FMT2

### Structural zeros
```{r struct_zero_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
FMT2_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment == "FMT2") %>% 
  dplyr::select(sample) %>%
  pull()

Donor_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment =="Donor") %>% 
  dplyr::select(sample) %>% pull()

sample_sub <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT2", "Donor"))

structural_zeros <- genome_counts_filt %>% 
  select(one_of(c("genome",subset_meta$sample))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>% 
   rowwise() %>% #compute for each row (genome)
   mutate(all_zeros_FMT2 = all(c_across(all_of(FMT2_samples)) == 0)) %>% 
   mutate(all_zeros_Donor = all(c_across(all_of(Donor_samples)) == 0)) %>% 
   mutate(average_FMT2 = mean(c_across(all_of(FMT2_samples)), na.rm = TRUE)) %>% 
   mutate(average_Donor = mean(c_across(all_of(Donor_samples)), na.rm = TRUE)) %>% 
   filter(all_zeros_FMT2 == TRUE || all_zeros_Donor==TRUE)  %>% 
   mutate(present = case_when(
      all_zeros_FMT2 & !all_zeros_Donor ~ "Donor",
      !all_zeros_FMT2 & all_zeros_Donor ~ "FMT2",
      !all_zeros_FMT2 & !all_zeros_Donor ~ "None",
      TRUE ~ NA_character_
    )) %>%
   mutate(average = ifelse(present == "FMT2", average_FMT2, average_Donor)) %>%
   dplyr::select(genome, present, average) %>%
   left_join(genome_metadata, by=join_by(genome==genome)) %>%
   arrange(present,-average)
```

***Structural zeros in each group***
```{r struc_count_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
fmt <- structural_zeros %>% 
  filter(present=="FMT2") %>% 
  count(phylum, name = "FMT2") %>%
  arrange(desc(FMT2)) 

fmt2_f <- structural_zeros %>% 
  filter(present=="FMT2") %>% 
  count(family, name = "FMT2") %>%
  arrange(desc(FMT2)) 
```

```{r total_sum_struc_tranp, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt, by="phylum" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%
  as.data.frame() %>% 
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))

```

***Phyla to which the structural zeros belong in each group***
```{r struc_count_merged_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt, by="phylum" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  tt()
```

***Families to which the structural zeros belong in each group***
```{r total_sum_family_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
structural_zeros %>% 
  filter(present=="Donor") %>% 
  count(family, name = "Donor") %>%
  arrange(desc(Donor)) %>% 
  full_join(.,fmt2_f, by="family" ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  tt()
```

#### Functional capacities of the structural zeros
```{r gift_struc_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% structural_zeros$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  filter(genome %in% structural_zeros$genome) %>% 
  select(one_of(c("genome",sample_sub$sample))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>% 
  column_to_rownames(., "genome")

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

```{r comunity_elem_FMT2_Tras, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c(9,10)], by=join_by("sample"=="sample"))
```

```{r commun_wilcox_elem_FMT2_trans, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  rownames_to_column(., "Elements")  %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))

element_gift_t <- element_gift  %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "trait")

element_gift_filt <- subset(element_gift_t, trait %in% significant_elements$trait) %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "sample")%>% 
  left_join(., sample_metadata[c(9,10)], by = join_by(sample == sample))

difference_table <- element_gift_filt %>%
  select(-sample) %>%
  group_by(treatment) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=FMT2-Donor)%>% 
  mutate(group_color = ifelse(Difference <0, "Donor","FMT2")) 
```

```{r commun_wilcox_elem_plot_fmt2_tranp, comment="", message=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.fullwidth=TRUE}
difference_table %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
  scale_fill_manual(
          breaks=c("FMT2","Donor"),
          values=c('#40714b50', "#d5b52c")) +
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Microbial Functional Capacity") + 
  ylab("Mean difference")
```

### Differential abundance analysis: composition
```{r phyloseq_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT2", "Donor")) %>% 
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>% 
  filter(!genome %in% structural_zeros$genome) %>% 
  select(one_of(c("genome",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
```

```{r ancom_rand_fmt2_tranp, comment="", message=FALSE, warning=FALSE}
ancombc_FMT2_Donor_2803 = ancombc2(data = physeq_genome_filtered, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "treatment",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = NULL, 
                  struc_zero = FALSE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

```{r save_ancombc_fmt2_tranp, warning=FALSE, comments="", message=FALSE}
save(ancombc_FMT2_Donor_2803, file = "data/ancombc_FMT2_Donor_2803.Rdata")
```

```{r load_ancombc_fmt2_tranp, warning=FALSE, comments="", message=FALSE, eval=FALSE}
load("data/ancombc_FMT2_Donor_2803.Rdata")
```

```{r ancom_rand, comment="", message=FALSE, warning=FALSE}
taxonomy <- data.frame(physeq_genome_filtered@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(order, phylum, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))

ancombc_rand_table_mag <- ancombc_FMT2_Donor_2803$res %>%
  dplyr::select(taxon, lfc_treatmentFMT2, p_treatmentFMT2) %>%
  filter(p_treatmentFMT2 < 0.05) %>%
  dplyr::arrange(p_treatmentFMT2) %>%
  merge(., taxonomy, by="taxon") %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_treatmentFMT2)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

***Differential abundance MAGs in each group***
```{r diff_fmt2_tran, comment="", message=FALSE, warning=FALSE}
fmt2_count <- ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT2<0)  %>% 
  count(phylum, name = "FMT2") %>%
  arrange(desc(FMT2)) 

ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT2>0)  %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor))  %>% 
  full_join(.,fmt2_count, by="phylum")%>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
```

```{r total_sum_fmt2_tran, comment="", message=FALSE, warning=FALSE}
ancombc_rand_table_mag %>%
  filter(lfc_treatmentFMT2>0)  %>% 
  count(phylum, name = "Donor") %>%
  arrange(desc(Donor))  %>% 
  full_join(.,fmt2_count, by="phylum")%>%
  mutate(across(where(is.numeric), ~ replace_na(., 0))) %>% 
  as.data.frame() %>% 
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))
```

```{r ancombc_rand_plot_phy_fmt2_tran, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag%>%
      mutate(genome=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
ggplot(., aes(x=lfc_treatmentFMT2, y=forcats::fct_reorder(genome,lfc_treatmentFMT2), fill=phylum)) +
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```
MAGs enriched in donor on the left side and in FMT2 on the right side



### Differences in the functional capacities
```{r gift_all_fmt2, comment="", message=FALSE, warning=FALSE}
sample_sub <- sample_metadata %>%
  filter(type == "Cold_intervention" & treatment %in% c("FMT2", "Donor"))

genome_counts_filtered <- genome_counts_filt %>% 
    select(one_of(c("genome",sample_sub$sample)))

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filtered$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_filtered_row <- genome_counts_filtered %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_filtered_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_filtered_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_filtered_row,GIFT_db)
```

#### Community elements differences:

***MCI***
```{r gitfs_elem_fmt2_tran, comment="", message=FALSE, warning=FALSE}
GIFTs_elements_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  summarise(MCI = mean(value), sd = sd(value))

MCI <- GIFTs_elements_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(MCI$value)
wilcox.test(value ~ treatment, data=MCI)
```
```{r comunity_elem_fmt2_tran, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c(9,10)], by=join_by("sample"=="sample"))
```

```{r commun_wilcox_elem_fmt2_tran, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  rownames_to_column(., "Elements")  %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))

element_gift_t <- element_gift  %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "trait")

element_gift_filt <- subset(element_gift_t, trait %in% significant_elements$trait) %>% 
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric)  %>%
  rownames_to_column(., "sample")%>% 
  left_join(., sample_metadata[c(9,10)], by = join_by(sample == sample))

difference_table <- element_gift_filt %>%
  select(-sample) %>%
  group_by(treatment) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=FMT2-Donor)%>% 
  mutate(group_color = ifelse(Difference <0, "Donor","FMT2")) 
```

#### Community functions differences

***MCI***
```{r gitfs_func_fmt2_tran, comment="", message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(treatment) %>%
  summarise(MCI = mean(value), sd = sd(value))

MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(MCI$value)
wilcox.test(value ~ treatment, data=MCI)
```
```{r comunity_func_fmt2_tran, comment="", message=FALSE, warning=FALSE}
function_gift <- GIFTs_functions_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  merge(., sample_metadata[c(9,10)], by="sample")
```

```{r commun_wilcox_func_fmt2_tran, comment="", message=FALSE, warning=FALSE}
unique_funct_db<- GIFT_db[c(3,4,5)] %>% 
  distinct(Code_function, .keep_all = TRUE)

significant_functional <- function_gift %>%
    pivot_longer(-c(sample,treatment), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ treatment)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)%>%
  left_join(.,unique_funct_db[c(1,3)],by = join_by(trait == Code_function))

significant_functional
```